---
title: Driving neurotabs from the Command Line
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Driving neurotabs from the Command Line}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
pkg_root <- if (file.exists("DESCRIPTION")) {
  "."
} else if (file.exists("../DESCRIPTION")) {
  ".."
} else {
  NULL
}
if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else if (!is.null(pkg_root) && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else if (requireNamespace("neurotabs", quietly = TRUE)) {
  library(neurotabs)
} else {
  stop("neurotabs must be installed, or pkgload/devtools must be available for local rendering.")
}

rscript_bin <- file.path(R.home("bin"), "Rscript")
roi_path <- system.file("examples/roi-only/nftab.yaml", package = "neurotabs")
stopifnot(nzchar(roi_path), file.exists(roi_path))
cli_script <- if (!is.null(pkg_root)) file.path(pkg_root, "exec", "neurotabs") else NULL
if (is.null(cli_script) || !file.exists(cli_script)) {
  stop("expected exec/neurotabs in the source tree for vignette rendering")
}

run_cli <- function(...) {
  args <- c(cli_script, ...)
  out <- system2(rscript_bin, args = args, stdout = TRUE, stderr = TRUE)
  status <- attr(out, "status")
  if (is.null(status)) status <- 0L
  list(status = status, output = out)
}

strip_rscript_noise <- function(lines) {
  lines[!grepl("^During startup|^[0-9]+: Setting LC_", lines)]
}

extract_json <- function(lines, simplifyVector = TRUE) {
  lines <- strip_rscript_noise(lines)
  start <- grep("^\\s*(\\{|\\[)", lines)[1L]
  end <- tail(grep("^\\s*(\\}|\\])\\s*$", lines), 1L)
  jsonlite::fromJSON(
    paste(lines[start:end], collapse = "\n"),
    simplifyVector = simplifyVector
  )
}
```

This vignette shows how to inspect, validate, materialize, and rewrite an
NFTab dataset from the shell. The goal is not to replace the R API. It is to
make the package usable in scripts, CI checks, and quick terminal workflows
where you want a stable command surface over the core `nf_*` functions.

The examples use the packaged ROI-only dataset, so every command is runnable as
written and every result can be checked in-code.

## What Does The CLI Expose?

The CLI wraps the same read, validate, resolve, collect, and write paths that
the package exposes in R.

```{r list-cli-commands}
help_out <- run_cli("help")
stopifnot(help_out$status == 0L)
cat(paste(utils::head(strip_rscript_noise(help_out$output), 15), collapse = "\n"))
```

The commands in this first pass are:

- `info`
- `validate`
- `features`
- `resolve`
- `collect`
- `copy`

## How Do You Inspect A Dataset Quickly?

Start with `info`. It reads the manifest, loads the observation table, and
prints a compact dataset summary.

```{r cli-info-json}
info_out <- run_cli("info", roi_path, "--json")
stopifnot(info_out$status == 0L)
info_payload <- extract_json(info_out$output)
info_payload
```

```{r cli-info-checks, include = FALSE}
stopifnot(
  identical(info_payload$dataset_id, "roi-only"),
  identical(info_payload$n_observations, 8L),
  identical(info_payload$features, "roi_beta"),
  identical(info_payload$axes, c("subject", "condition"))
)
```

That gives you the minimum orientation you need before doing anything heavier:
which dataset you are reading, how many rows it has, which features it
declares, and which observation axes define row identity.

## How Do You Check Conformance?

The `validate` command runs the same structural or full conformance checks that
`nf_validate()` provides in R.

```{r cli-validate-structural}
validate_out <- run_cli("validate", roi_path, "--level", "structural", "--json")
stopifnot(validate_out$status == 0L)
validate_payload <- extract_json(validate_out$output)
validate_payload
```

```{r cli-validate-checks, include = FALSE}
stopifnot(
  isTRUE(validate_payload$valid),
  identical(validate_payload$level, "structural"),
  length(validate_payload$errors) == 0L
)
```

If you want the stronger contract, use `--level full`. That attempts to resolve
every non-missing feature value, so it is the better fit for CI or release
checks.

```{r cli-validate-full}
full_out <- run_cli("validate", roi_path, "--level", "full", "--json")
stopifnot(full_out$status == 0L)
full_payload <- extract_json(full_out$output)
full_payload[c("valid", "level", "dataset_id")]
```

## How Do You Discover The Feature Surface?

Use `features` when you need to know what a dataset can materialize without
opening the manifest by hand.

```{r cli-features}
features_out <- run_cli("features", roi_path, "--json")
stopifnot(features_out$status == 0L)
features_payload <- extract_json(features_out$output, simplifyVector = FALSE)
features_payload
```

```{r cli-features-checks, include = FALSE}
stopifnot(
  length(features_payload) == 1L,
  identical(features_payload[[1]]$name, "roi_beta"),
  identical(features_payload[[1]]$kind, "vector"),
  identical(features_payload[[1]]$axes, "roi")
)
```

For command-line work, this is the bridge between “there is a manifest here”
and “I know which feature names are valid arguments to `resolve` or `collect`”.

## How Do You Materialize One Row?

`resolve` is the CLI equivalent of `nf_resolve()`. You supply a feature plus a
row selector, either by `row_id` or by 1-based row index.

```{r cli-resolve}
resolve_out <- run_cli("resolve", roi_path, "roi_beta", "--row", "r01")
stopifnot(resolve_out$status == 0L)
resolve_payload <- extract_json(resolve_out$output)
resolve_payload
```

```{r cli-resolve-checks, include = FALSE}
stopifnot(
  identical(resolve_payload$row_id, "r01"),
  length(resolve_payload$value) == 5L,
  all(is.finite(resolve_payload$value)),
  abs(resolve_payload$value[[1]] - 0.31) < 1e-6
)
```

This is the right tool when you are debugging one bad row, inspecting a single
feature binding, or spot-checking data before a larger job.

## How Do You Materialize A Feature Across All Rows?

`collect` resolves one feature across the full dataset. For fixed-size 1D
features, it can emit a tabular result directly.

```{r cli-collect}
collect_file <- tempfile(fileext = ".csv")
collect_status <- run_cli("collect", roi_path, "roi_beta", "--out", collect_file, "--format", "csv")
stopifnot(collect_status$status == 0L, file.exists(collect_file))
collect_tbl <- utils::read.csv(collect_file, stringsAsFactors = FALSE)
collect_tbl
```

```{r cli-collect-checks, include = FALSE}
stopifnot(
  nrow(collect_tbl) == 8L,
  ncol(collect_tbl) == 6L,
  identical(collect_tbl$row_id[[1]], "r01"),
  all(vapply(collect_tbl[-1], is.numeric, logical(1L))),
  all(is.finite(as.matrix(collect_tbl[-1])))
)
```

That output is intentionally simple: one row per observation, one leading
`row_id` column, and one column per resolved element.

## How Do You Rewrite A Dataset Cleanly?

`copy` is a read-and-write round trip. It is useful when you want to normalize
an existing dataset layout, rewrite checksums, or stage a portable copy for
another tool.

```{r cli-copy}
copy_dir <- tempfile("neurotabs-cli-copy-")
copy_out <- run_cli("copy", roi_path, copy_dir, "--json")
stopifnot(copy_out$status == 0L)
copy_payload <- extract_json(copy_out$output)
copy_payload
```

```{r cli-copy-checks, include = FALSE}
copied_ds <- nf_read(copy_payload$manifest)
stopifnot(
  identical(copied_ds$manifest$dataset_id, "roi-only"),
  identical(nf_nobs(copied_ds), 8L),
  identical(nf_feature_names(copied_ds), "roi_beta")
)
```

Because the command goes through `nf_read()` and `nf_write()`, it follows the
same manifest parsing, dtype coercion, and checksum-writing rules as the R API.

## When Should You Use The CLI Versus The R API?

Use the CLI when you need:

- a stable shell command in CI or Make targets
- a fast conformance check before handing data to another tool
- a quick summary or one-off materialization without writing R code

Use the R API when you need:

- filtering, grouping, and compute verbs
- integration with downstream analysis code
- custom control flow around errors, grouping, or feature transformations

The command-line layer is deliberately small. It covers the package’s core
dataset lifecycle without trying to reproduce the full interactive grammar in
shell flags.

## What Should You Read Next?

For the end-to-end R workflow, start with `vignette("neurotabs")`. For the
manifest contract itself, use `vignette("specification")`.
