---
title: Getting Started with neurotabs
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with neurotabs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
pkg_root <- if (file.exists("DESCRIPTION")) {
  "."
} else if (file.exists("../DESCRIPTION")) {
  ".."
} else {
  NULL
}
if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else if (!is.null(pkg_root) && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else if (requireNamespace("neurotabs", quietly = TRUE)) {
  library(neurotabs)
} else {
  stop("neurotabs must be installed, or pkgload/devtools must be available for local rendering.")
}
```

`neurotabs` is the reference R implementation of the NFTab specification. You
use it to read an NFTab manifest, validate the dataset contract, and resolve
feature values without hard-coding storage details into your analysis.

## What Problem Does neurotabs Solve?

Neuroimaging projects often end up as a directory of map files plus one or more
spreadsheets that describe subjects, conditions, runs, and derived features.
That works until you need to answer questions like:

- Which rows belong to the patient cohort?
- Which file or selector encodes the statistical map for this row?
- Are two study outputs compatible enough to concatenate safely?
- Can another tool read the dataset without re-implementing your file-discovery logic?

`neurotabs` addresses that problem by separating the logical dataset contract
from the physical storage. An NFTab manifest says what each row means, what
each feature is, and how a feature value resolves for that row.

## When Should You Use It?

The package is most useful when your dataset is row-oriented and each row has
both design metadata and one or more neuroimaging features.

Typical Phase 1 use cases are:

1. Read a collection of ROI vectors or statistical maps together with subject
   and condition metadata.
2. Validate that the dataset is structurally well-formed before analysis or
   handoff to another tool.
3. Resolve one feature across many rows without hard-coding column bundles,
   file paths, or selector logic.
4. Concatenate compatible datasets from different studies or storage layouts.

## What Does An NFTab Dataset Look Like?

The package ships with a small ROI-only example that uses `columns` encoding.

```{r load-roi-example}
roi_path <- system.file("examples/roi-only/nftab.yaml", package = "neurotabs")
stopifnot(nzchar(roi_path))

roi_ds <- nf_read(roi_path)
roi_ds
```

An `nftab` object keeps the parsed manifest, the observation table, and any
optional resource registry together. The manifest defines the observation axes
and the feature schemas:

```{r inspect-roi-example}
nf_axes(roi_ds)
nf_feature_names(roi_ds)
```

## What Happens When You Read A Dataset?

`nf_read()` does more than parse YAML. It validates the manifest against the
bundled JSON Schema, reads the observation table, and coerces columns to the
declared NFTab dtypes.

```{r inspect-read-types}
vapply(
  roi_ds$observations[c("row_id", "subject", "age", "roi_1")],
  function(x) paste(class(x), collapse = "/"),
  character(1)
)
```

That type coercion is part of the package contract. If a manifest says a column
is `float64` or `date`, `nf_read()` normalizes the loaded table to match.

## How Do You Check Conformance?

Structural conformance answers the question “is this dataset well-formed?”.
Full conformance goes further and tries to resolve every non-missing feature.

```{r validate-roi-example}
structural <- nf_validate(roi_ds, level = "structural")
stopifnot(structural$valid)

structural$valid
```

For the ROI-only example, full conformance is also straightforward because every
feature value lives directly in observation-table columns.

```{r validate-roi-full}
full <- nf_validate(roi_ds, level = "full")
stopifnot(full$valid)

full$valid
```

## How Do Feature Values Resolve?

The ROI example stores one logical feature, `roi_beta`, as an ordered set of
scalar columns. You can resolve one row at a time or collect the feature across
all rows.

```{r resolve-roi}
nf_resolve(roi_ds, 1L, "roi_beta")
```

```{r collect-roi}
roi_mat <- nf_collect(roi_ds, "roi_beta")
stopifnot(is.matrix(roi_mat), ncol(roi_mat) == 5L)

dim(roi_mat)
```

The package also ships a mixed example with both `columns` and `ref` encodings.
The ROI feature still resolves immediately even though the dataset also declares
a volume feature.

```{r load-faces-example}
faces_path <- system.file("examples/faces-demo/nftab.yaml", package = "neurotabs")
stopifnot(nzchar(faces_path))

faces_ds <- nf_read(faces_path)
nf_feature_names(faces_ds)
nf_resolve(faces_ds, 1L, "roi_beta")
```

## How Do You Encode ROI Names And Atlas Identity?

For `columns` encodings, the physical column names are just storage. The
semantic metadata belongs to the feature axis. In practice that means:

- `binding.columns` defines the storage order of the ROI vector.
- `nf_axis_info(..., "roi")$id` identifies the atlas or parcellation.
- `nf_axis_info(..., "roi")$labels` points to a label table that names each ROI.

The shipped `faces-demo` uses that pattern for its three-region ROI feature.

```{r inspect-axis-domain}
roi_schema <- nf_feature_schema(faces_ds, "roi_beta")
stopifnot(identical(roi_schema$axes, "roi"))

roi_axis <- nf_axis_info(faces_ds, "roi_beta", "roi")
stopifnot(identical(roi_axis$id, "desikan3-demo"))

roi_axis
```

```{r read-axis-labels}
roi_labels <- nf_axis_labels(faces_ds, "roi_beta", "roi")
roi_values <- nf_resolve(faces_ds, 1L, "roi_beta")
stopifnot(nrow(roi_labels) == length(roi_values))

data.frame(
  atlas = roi_axis$id,
  label = roi_labels$label,
  value = roi_values,
  row.names = NULL
)
```

This is the intended encoding for parcel-wise vectors. The atlas identity lives
with the logical axis, not in the storage column names, so the same logical ROI
feature can be stored under different physical column bundles and still mean the
same thing.

## What Does Full Conformance Mean For External Resources?

For `ref` encodings, full conformance requires successful materialization
through a registered backend. The package also validates resource checksums
before backend dispatch when a checksum is present.

```{r checksum-demo-setup, include = FALSE}
tmpdir <- tempfile("neurotabs-vignette-")
dir.create(tmpdir)

payload_path <- file.path(tmpdir, "payload.bin")
writeBin(as.raw(c(1, 2, 3, 4)), payload_path)

obs_cols <- list(
  row_id = nf_col_schema("string", nullable = FALSE),
  subject = nf_col_schema("string", nullable = FALSE),
  condition = nf_col_schema("string", nullable = FALSE),
  res_id = nf_col_schema("string", nullable = FALSE)
)

feature <- nf_feature(
  logical = nf_logical_schema("volume", c("x", "y", "z"), "float32",
                              shape = c(1L, 1L, 1L), support_ref = "demo_grid"),
  encodings = list(nf_ref_encoding(resource_id = nf_col("res_id")))
)

manifest <- nf_manifest(
  dataset_id = "checksum-demo",
  row_id = "row_id",
  observation_axes = c("subject", "condition"),
  observation_columns = obs_cols,
  features = list(statmap = feature),
  supports = list(
    demo_grid = nf_support_volume(
      support_id = "demo-grid-1x1x1",
      space = "MNI152NLin2009cAsym",
      grid_id = "demo-grid-1x1x1"
    )
  ),
  resources_path = "resources.csv",
  resources_format = "csv"
)

observations <- data.frame(
  row_id = "r1",
  subject = "s1",
  condition = "faces",
  res_id = "res1",
  stringsAsFactors = FALSE
)

resources <- data.frame(
  resource_id = "res1",
  backend = "x-unknown",
  locator = payload_path,
  checksum = "md5:00000000000000000000000000000000",
  stringsAsFactors = FALSE
)

checksum_ds <- nftab(manifest, observations, resources = resources, .root = tmpdir)
checksum_error <- try(nf_resolve(checksum_ds, 1L, "statmap"), silent = TRUE)
checksum_detected <- inherits(checksum_error, "try-error")
```

```{r checksum-demo}
stopifnot(isTRUE(checksum_detected))

checksum_detected
```

A `TRUE` result here means the resolver rejected the resource before it reached
backend dispatch because the checksum token did not match the file on disk.

## How Do You Create a Dataset?

### The Quick Way: `nf_from_table()`

If you have a table (CSV or data.frame) and a set of NIfTI files, one per row:

```{r from-table-3d, eval = FALSE}
ds <- nf_from_table(
  "participants.csv",          # table with subject, condition, etc.
  feature     = "statmap",     # name for the feature
  locator_col = "nifti_path",  # column pointing to .nii.gz files
  space       = "MNI152NLin2009cAsym",
  root        = "/data/study"
)
```

Or if all rows map to successive volumes in a single 4D NIfTI:

```{r from-table-4d, eval = FALSE}
ds <- nf_from_table(
  design_table,                      # data.frame with subject, condition, etc.
  feature = "bold",
  locator = "results/group_stats.nii.gz",  # shared 4D file
  space   = "MNI152NLin2009cAsym",
  root    = "/data/study"
)
# Row 1 -> volume 0, row 2 -> volume 1, etc.
```

### The Manual Way: Full Control

Every observation column needs a declared dtype and nullability:

```{r create-columns}
obs_cols <- list(
  row_id    = nf_col_schema("string",  nullable = FALSE),
  subject   = nf_col_schema("string",  nullable = FALSE),
  condition = nf_col_schema("string",  nullable = FALSE),
  roi_1     = nf_col_schema("float32"),
  roi_2     = nf_col_schema("float32")
)
```

### Step 2: Define Features

A feature has a logical schema (what the resolved value IS) and one or more
encodings (how to find it). For a 2-element ROI vector stored in columns:

```{r create-feature}
roi_feature <- nf_feature(
  logical = nf_logical_schema(
    kind  = "vector",
    axes  = "roi",
    dtype = "float32",
    shape = 2L
  ),
  encodings = list(
    nf_columns_encoding(c("roi_1", "roi_2"))
  )
)
```

### Step 3: Build the Manifest and Dataset

The manifest ties columns, features, and axes together:

```{r create-manifest}
manifest <- nf_manifest(
  dataset_id          = "my-study",
  row_id              = "row_id",
  observation_axes    = c("subject", "condition"),
  observation_columns = obs_cols,
  features            = list(roi = roi_feature)
)
```

Then combine the manifest with an observation table:

```{r create-nftab}
observations <- data.frame(
  row_id    = c("r1", "r2", "r3", "r4"),
  subject   = c("s01", "s01", "s02", "s02"),
  condition = c("faces", "houses", "faces", "houses"),
  roi_1     = c(0.5, 0.2, 0.6, 0.3),
  roi_2     = c(0.4, 0.1, 0.5, 0.2),
  stringsAsFactors = FALSE
)

my_ds <- nftab(manifest, observations)
my_ds
```

### Step 4: Write to Disk

`nf_write()` produces a portable NFTab directory with a YAML manifest and CSV
observation table:

```{r write-dataset}
out_dir <- tempfile("my-study-")
nf_write(my_ds, out_dir)

list.files(out_dir)
```

The written dataset can be read back with `nf_read()`:

```{r read-back}
my_ds2 <- nf_read(file.path(out_dir, "nftab.yaml"))
nf_resolve(my_ds2, "r1", "roi")
```

## How Do You Filter, Group, and Summarize?

`neurotabs` provides dplyr-style grammar verbs that operate directly on `nftab`
objects. These verbs work with both columns-encoded and NIfTI-backed features.

### Filtering Rows

`nf_filter()` selects observations matching a predicate. Observation column
names are available directly in the expression:

```{r filter-example}
ctrl_ds <- nf_filter(roi_ds, group == "ctrl")
nf_nobs(ctrl_ds)
```

### Selecting and Arranging

`nf_select()` keeps named design columns (encoding-required columns are always
retained). `nf_arrange()` sorts rows:

```{r select-arrange}
sorted <- nf_arrange(roi_ds, subject)
nf_design(sorted)[, c("row_id", "subject", "condition")]
```

### Applying Functions Across Rows

`nf_apply()` runs a function or fixed operation on each row's feature value.
For character operations like `"mean"`, `"sum"`, `"sd"`, the package uses
optimized batch paths:

```{r apply-example}
means <- nf_apply(roi_ds, "roi_beta", "mean")
means
```

### Grouping and Summarizing

`nf_group_by()` + `nf_summarize()` computes group-level feature summaries.
The result is a new `nftab` with one row per group:

```{r group-summarize}
grouped <- nf_group_by(roi_ds, group)
summary_ds <- nf_summarize(grouped, "roi_beta", .f = "mean")
summary_ds
```

You can resolve the summarized feature to see the group-level average:

```{r resolve-summary}
nf_resolve(summary_ds, 1L, "roi_beta")
```

### Adding Derived Columns

`nf_mutate()` adds new observation columns. Inside `nf_mutate()`, the helper
`nf_apply_feature()` derives scalar values from features:

```{r mutate-example}
ds_with_mean <- nf_mutate(roi_ds, roi_mean = nf_apply_feature("roi_beta", "mean"))
ds_with_mean$observations[, c("row_id", "subject", "roi_mean")]
```

## Where Should You Go Next?

The full workflow is:

1. Read a manifest with `nf_read()`.
2. Check structural or full conformance with `nf_validate()`.
3. Filter, arrange, and select with `nf_filter()`, `nf_arrange()`, `nf_select()`.
4. Resolve features with `nf_resolve()` or `nf_collect()`.
5. Apply row-wise operations with `nf_apply()`.
6. Group and summarize with `nf_group_by()` + `nf_summarize()`.
7. Compare groups with `nf_compare()`.
8. Combine compatible datasets with `nf_concat()`.

For NIfTI-backed features, set `options(neurotabs.compute.workers = 4L)` to
enable parallel file reads on multi-core systems.
