---
title: Understanding the NFTab Specification
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding the NFTab Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
pkg_root <- if (file.exists("DESCRIPTION")) {
  "."
} else if (file.exists("../DESCRIPTION")) {
  ".."
} else {
  NULL
}
if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else if (!is.null(pkg_root) && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
} else if (requireNamespace("neurotabs", quietly = TRUE)) {
  library(neurotabs)
} else {
  stop("neurotabs must be installed, or pkgload/devtools must be available for local rendering.")
}
```

This vignette is a guided reading of the NFTab specification. The full
normative text lives in `spec/nftab-spec.md` in the source repository; this
document explains the same contract in package terms and uses the shipped
example datasets to make the main ideas concrete.

If you want the end-to-end analysis workflow first, start with
`vignette("neurotabs")`. This article assumes you already know how to read and
query a dataset and focuses instead on what the manifest contract means, how
logical features stay separate from storage, and why compatibility is
schema-driven.

## What Does The Spec Standardize?

NFTab standardizes a row-oriented neuroimaging dataset with three parts:

- a manifest that defines the logical contract
- an observation table that stores one row per observation
- an optional resource registry for externally stored feature data

The examples shipped with `neurotabs` show both the simple and mixed cases. In
this vignette, the goal is not to repeat the grammar verbs. It is to compare
raw manifest fragments with the parsed dataset objects and inspect the contract
they enforce.

```{r load-spec-examples}
roi_path <- system.file("examples/roi-only/nftab.yaml", package = "neurotabs")
faces_path <- system.file("examples/faces-demo/nftab.yaml", package = "neurotabs")
stopifnot(nzchar(roi_path), nzchar(faces_path))

roi_raw <- yaml::read_yaml(roi_path)
faces_raw <- yaml::read_yaml(faces_path)

roi_ds <- nf_read(roi_path)
faces_ds <- nf_read(faces_path)
```

```{r abstract-model}
list(
  dataset_id = roi_raw$dataset_id,
  storage_profile = roi_raw$storage_profile,
  observation_axes = roi_raw$observation_axes,
  features = names(roi_raw$features)
)
```

That output corresponds directly to the abstract data model in the spec: rows
are observations, features resolve per row, and the manifest defines what those
resolved values mean independent of physical storage.

## What Does The Manifest Declare?

At the top level, a manifest declares dataset identity, storage profile, row
identity, observation axes, scalar observation columns, and the feature set.

```{r manifest-top-level}
roi_raw[c(
  "spec_version",
  "dataset_id",
  "storage_profile",
  "row_id",
  "observation_axes"
)]
```

The observation table location is part of the manifest too. In the
`table-package` storage profile, the table is a CSV, TSV, or Parquet file that
travels with the manifest.

```{r observation-table-ref}
roi_raw$observation_table
```

## What Is An Observation Table?

The observation table is the row-oriented part of the dataset. Each row is one
observation, and scalar design metadata such as subject, group, condition, or
age live directly in table columns.

```{r observation-table-preview}
utils::head(
  roi_ds$observations[c("row_id", "subject", "group", "condition", "age")],
  4
)
```

Each observation column has a declared scalar schema. The spec uses that schema
to define dtype, nullability, semantic role, levels, and optional units.

```{r scalar-schema}
roi_raw$observation_columns$age
```

When `nf_read()` loads the dataset, it coerces table columns to those declared
types and rejects incompatible values. The getting-started vignette shows how
you work with those columns interactively; here the important point is that the
schema is part of the portable contract.

## What Is A Feature Schema?

A feature is not just a storage column or file path. In NFTab, a feature has a
logical schema that defines the kind of value produced when the feature is
resolved for one row.

```{r logical-schema}
roi_schema <- nf_feature_schema(roi_ds, "roi_beta")
roi_schema
```

For the ROI example, `roi_beta` is a `vector` with one semantic axis called
`roi`, element dtype `float32`, and logical shape `5`. Resolving one row
produces that logical value.

```{r logical-value}
roi_value <- nf_resolve(roi_ds, 1L, "roi_beta")
stopifnot(length(roi_value) == 5L)

roi_value
```

This separation between logical schema and physical storage is the center of
the spec. It is what lets two datasets remain compatible even when they use
different physical layouts, as long as the resolved feature contract is the
same.

## How Do Encodings Work?

The spec allows a feature to have one or more encodings. An encoding says how a
logical feature value can be materialized for one row.

The ROI example uses a `columns` encoding: the logical vector is stored in an
ordered bundle of scalar table columns.

```{r columns-encoding}
roi_raw$features$roi_beta$encodings[[1]]
```

The mixed `faces-demo` uses a `ref` encoding for a volume feature. Instead of
putting voxel data in the observation table, each row carries a resource
identifier and an optional selector.

```{r ref-encoding}
faces_raw$features$statmap$encodings[[1]]
```

This is why the spec distinguishes between logical features and storage
backends. The feature means “a 3D statistical map in MNI space”, while the
encoding says how to find it for a particular row.

## How Do Axis Domains Carry Atlas Metadata?

Axis domains are where the spec puts semantic metadata for a feature axis. For
parcel-wise vectors, this is how you encode atlas identity and ROI labels.

```{r axis-domain}
roi_axis <- nf_axis_info(faces_ds, "roi_beta", "roi")
roi_labels <- nf_axis_labels(faces_ds, "roi_beta", "roi")
stopifnot(nrow(roi_labels) == 3L)

roi_axis
```

```{r axis-labels}
data.frame(
  atlas = roi_axis$id,
  label = roi_labels$label,
  row.names = NULL
)
```

This matches the intended reading of the spec:

- the physical `binding.columns` order defines storage order
- the `roi` axis defines what those positions mean
- the axis domain names the atlas and supplies the label table

## How Do Sparse Masked Volumes Fit?

Some datasets do not store a dense 3D volume for each row. Instead, they store
only the values inside a fixed brain mask. In that case the logical feature is
better modeled as a vector over a `voxel` axis, together with an extension that
maps vector positions back to grid coordinates.

```{r masked-volume-extension}
masked_manifest <- nf_manifest(
  dataset_id = "masked-demo",
  row_id = "row_id",
  observation_axes = c("subject", "condition"),
  observation_columns = list(
    row_id = nf_col_schema("string", nullable = FALSE),
    subject = nf_col_schema("string", nullable = FALSE),
    condition = nf_col_schema("string", nullable = FALSE),
    v1 = nf_col_schema("float32"),
    v2 = nf_col_schema("float32"),
    v3 = nf_col_schema("float32")
  ),
  features = list(
    statvec = nf_feature(
      logical = nf_logical_schema(
        "vector",
        "voxel",
        "float32",
        shape = 3L,
        space = "MNI152NLin2009cAsym",
        alignment = "same_grid"
      ),
      encodings = list(nf_columns_encoding(c("v1", "v2", "v3")))
    )
  ),
  extensions = list(
    "x-masked-volume" = list(
      features = list(
        statvec = list(
          grid_axes = c("x", "y", "z"),
          grid_shape = c(91L, 109L, 91L),
          grid_index_base = 0L,
          index_map = list(
            path = "voxel_index.tsv",
            format = "tsv"
          )
        )
      )
    )
  )
)

masked_ds <- nftab(
  masked_manifest,
  observations = data.frame(
    row_id = c("r1", "r2"),
    subject = c("s01", "s02"),
    condition = c("faces", "houses"),
    v1 = c(0.1, 0.4),
    v2 = c(0.2, 0.5),
    v3 = c(0.3, 0.6),
    stringsAsFactors = FALSE
  )
)

nf_extension(masked_ds, "x-masked-volume")
```

That extension says the feature is still stored as a vector, but the manifest
also supplies an external index map telling readers which `(x, y, z)` voxel
each vector position refers to. This is the common case for masked group maps
or sparse voxelwise features.

## What Is The Resource Registry For?

A `ref` encoding can point directly to a backend and locator, but many datasets
use a resource registry so rows can refer to shared resources by `resource_id`.

```{r resource-registry-manifest}
faces_raw$resources
```

```{r resource-registry-table}
faces_ds$resources
```

The spec requires that registry rows supply the storage metadata needed to
materialize the referenced resource, such as backend name, locator, and
optionally checksum.

## What Does Conformance Mean?

The spec distinguishes between structural and full conformance.

- Structural conformance asks whether the manifest, table, and declared schemas
  form a well-typed NFTab dataset.
- Full conformance goes further and attempts to resolve every non-missing
  feature value.

For the ROI-only dataset, both levels pass because the feature values live in
table columns and require no external backend.

```{r conformance}
structural <- nf_validate(roi_ds, level = "structural")
full <- nf_validate(roi_ds, level = "full")
stopifnot(structural$valid, full$valid)

data.frame(
  level = c("structural", "full"),
  valid = c(structural$valid, full$valid),
  row.names = NULL
)
```

For datasets with `ref` encodings, full conformance additionally depends on
backend support and resource integrity checks such as checksum validation.

## How Does Schema-Driven Compatibility Work?

The spec also defines when two datasets are compatible enough to concatenate.
That decision depends on matching observation axes, compatible scalar columns,
and identical logical feature schemas.

```{r compatibility-summary}
compat_roi <- nf_compatible(roi_ds, roi_ds)
compat_faces <- nf_compatible(faces_ds, faces_ds)
compat_cross <- nf_compatible(roi_ds, faces_ds)

compat_df <- data.frame(
  comparison = c("ROI vs ROI", "Faces vs Faces", "ROI vs Faces"),
  compatible = c(
    compat_roi$compatible,
    compat_faces$compatible,
    compat_cross$compatible
  ),
  mismatches = c(
    length(compat_roi$reasons),
    length(compat_faces$reasons),
    length(compat_cross$reasons)
  ),
  row.names = NULL
)

compat_df
```

```{r compatibility-plot, echo = FALSE, fig.width = 6, fig.height = 3.5}
plot_df <- transform(
  compat_df,
  comparison = factor(comparison, levels = rev(comparison)),
  status = ifelse(compatible, "compatible", "incompatible"),
  label = ifelse(compatible, "exact match", paste(mismatches, "mismatches"))
)

ggplot2::ggplot(plot_df, ggplot2::aes(x = "schema check", y = comparison, fill = status)) +
  ggplot2::geom_tile(width = 0.85, height = 0.8, color = "white") +
  ggplot2::geom_text(ggplot2::aes(label = label), size = 3.5) +
  ggplot2::scale_fill_manual(values = c(compatible = "#3b7a57", incompatible = "#c8553d")) +
  ggplot2::labs(x = NULL, y = NULL, fill = NULL) +
  ggplot2::theme(
    legend.position = "top",
    panel.grid = ggplot2::element_blank()
  )
```

The compatible pairs pass because their observation axes, scalar schemas, and
logical feature contracts line up exactly. The mixed pair fails for schema
reasons, not because the files happen to live in different places.

```{r compatibility-reasons}
compat_cross$reasons
```

In other words, NFTab concatenation is schema-driven. It is not just row-binding
two tables and hoping the feature values still mean the same thing.

## How Should You Read The Full Spec?

The most important sections to understand first are:

1. Abstract data model
2. Observation table and scalar column schema
3. Feature schema and logical schema
4. Encodings and resource registry
5. Resolution, conformance, and compatibility

If you keep those five layers in mind, the rest of the specification reads as a
precise contract rather than a collection of disconnected fields.
